\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{enumitem}

\usepackage[margin=0.7in]{geometry} % Adjust the margins of document here.

\usepackage{tikz}                                          % Для простых рисунков в документе
\usetikzlibrary{matrix,arrows,decorations.pathmorphing,shapes.geometric,calc,snakes,backgrounds,arrows.meta}
\usepackage{xcolor}

\begin{document}
\pagestyle{plain}

\section*{Problems from IMC}

\textbf{First day — July 30, 2011}

\textbf{Problem 1.} (10 points)
Let \( f : \mathbb{R} \to \mathbb{R} \) be a continuous function.
A point \( x \) is called a \textit{shadow point} if there exists a point \( y \in \mathbb{R} \)
with \( y > x \) such that \( f(y) > f(x) \). Let \( a < b \) be real numbers and suppose that
\begin{itemize}
    \item all the points of the open interval \( I = (a, b) \) are shadow points;
    \item \( a \) and \( b \) are not shadow points.
\end{itemize}
Prove that
\begin{enumerate}
    \item \( f(x) \leq f(b) \) for all \( a < x < b \);
    \item \( f(a) = f(b) \).
\end{enumerate}

\textbf{Problem 2.} (10 points)
Does there exist a real \( 3 \times 3 \) matrix \( A \) such that \( \text{tr}(A) = 0 \) and \( A^2 + A^t = I \)?
(Here, \(\text{tr}(A)\) denotes the trace of \( A \), \( A^t \) is the transpose of \( A \), and \( I \) is the identity matrix.)

\textbf{Problem 3.} (10 points)
Let \( p \) be a prime number. Call a positive integer \( n \) \textit{interesting} if
\[
x^n - 1 = (x^p - x + 1) f(x) + p g(x)
\]
for some polynomials \( f \) and \( g \) with integer coefficients.
\begin{enumerate}
    \item Prove that the number \( p^p - 1 \) is interesting.
    \item For which \( p \) is \( p^p - 1 \) the minimal interesting number?
\end{enumerate}

\textbf{Problem 4.} (10 points)
Let \( A_1, A_2, \ldots, A_n \) be finite, nonempty sets. Define the function
\[
f(t) = \sum_{k=1}^{n} \sum_{1 \leq i_1 < i_2 < \cdots < i_k \leq n} (-1)^{k-1} t^{|A_{i_1} \cup A_{i_2} \cup \cdots \cup A_{i_k}|}.
\]
Prove that \( f \) is nondecreasing on \([0, 1]\).

(\(|A|\) denotes the number of elements in \(A\).)

\textbf{Problem 5.} (10 points)
Let \( n \) be a positive integer and let \( V \) be a \( (2n-1) \)-dimensional vector space over the two-element field. Prove that for arbitrary vectors \( v_1, \ldots, v_{4n-1} \in V \), there exists a sequence \( 1 \leq i_1 < \cdots < i_{2n} \leq 4n-1 \) of indices such that
\[
v_{i_1} + \cdots + v_{i_{2n}} = 0.
\]

\end{document}
